Nested sampling with two objective functions

### 2016-02-10 Paper/ms.tex
ht!
Some notational fixes

$Z$ is useful because it is the probability of the data given that
$\xx$ is {\em one of}
the possibilities in the space being integrated over, and thus plays the role
of a likelihood for the parameter space as a whole.
The main goal of NS is to approximate $Z$, but it can also be used to make
Monte Carlo approximations of the posterior distribution, and to
calculate the `information',
or Kullback-Leibler divergence from the prior to the posterior:
\begin{eqnarray}
\mathcal{H} &=& \int p(\xx | \data, \II) \ln
\left[\frac{p(\xx | \data, \II)}{p(\xx | \II)}\right] \, d\xx \\
&=& \int \frac{\pi(\xx) L(\xx)}{Z} \ln
\left[\frac{L(\xx)}{Z}\right] \, d\xx.
\end{eqnarray}
$\mathcal{H}$ quantifies how compressed the posterior distribution is with
respect to the prior. For example, $\mathcal{H} = 10$ implies (loosely speaking)
that the posterior occupies a fraction $e^{-10}$ of the prior mass.
$\mathcal{H}$ can be interpreted quite literally as a measure of how much
the data $\data$ told you about $\xx$.

@@

The prior $\pi(\xx)$ implies a certain prior for $L_1$ and $L_2$, which we
denote $\pi(L_1, L_2)$. An example of a prior is shown as the
probability density in Figure~\ref{fig:joint1}. The partition function
$Z(\beta_1, \beta_2)$ is a set of expected values with respect to this density.
Simple Monte Carlo sampling from $\pi$ will not work except for values of
$(\beta_1, \beta_2)$ where the canonical distribution remains similar to $\pi$.
We need a sampler that explores regions where $L_1$ and $L_2$ are high in order
to accurately estimate $Z(\beta_1, \beta_2)$.
\begin{figure}
\centering
\includegraphics[scale=0.5]{figures/joint1.pdf}
\caption{\it An example of a prior for $L_1$ and $L_2$, which is implied by
the prior $\pi(\xx)$. We have overplotted 25 points drawn from this prior.
The estimated prior mass in the shaded rectangle is
$\hat{X}(2, 1.5) = 5/25 = 0.2$.
\label{fig:joint1}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.5]{figures/contours.pdf}
\caption{\it The joint prior for the objective functions, and fifteen
contours of the cumulative distribution $X(L_1, L_2)$, each enclosing half
the prior mass of the previous one. \label{fig:contours}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.5]{figures/joint2.pdf}
\caption{\it The joint prior for the objective functions, and Monte Carlo
samples, as in Figure~\ref{fig:joint1}. The shaded region imposes a
constraint on the value of $\hat{X}$. Its boundary approximates a contour
of $X(L_1, L_2)$.
\label{fig:joint2}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.75]{figures/truth.pdf}
\caption{The true partition function and KL divergence (from the prior to
the canonical distribution) as a function of two temperatures for the
test problem.\label{fig:truth}}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=1]{figures/output.png}
\caption{The sequence of discarded particles (yellow) and
points representing the canonical distribution at temperatures
$(T_1, T_2) = (0.1, 1)$ (black).\label{fig:output}}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=0.5]{figures/results.pdf}
\caption{Numerical results for the test problem. These compare well with the
true partition function and KL divergence given in Figure~\ref{fig:truth}.
\label{fig:results}}
\end{figure}

@@

\section{Example 3: ``MaxEnt'' image deconvolution example}
Problems with more than one objective function can also arise in data
analysis. Consider, for example, the problem of deblurring a noisy
image. A common approach to these problems is regularized inversion,
such as the ``maximum entropy'' technique of \citet{gull}. It is important
to distinguish between this method and the maximum entropy principle used
to update or assign probabilities \citep{caticha}.

The unknown parameters are a set of pixel intensities
$\{m_i\}_{i=1}^n$ where $n$ is the number of pixels and $m_i \geq 0$.
The dataset is also a collection of pixel intensities $\{D_i\}_{i=1}^n$.
The prior for the dataset given the parameters is a normal distribution
with mean given by the underlying image $\{m_i\}$ convolved by
a point spread function. The standard deviation of each pixel is
$\sigma$, assumed known. This is usually written as

\begin{eqnarray}
\boldsymbol{D} = \boldsymbol{P}\boldsymbol{m} + \boldsymbol{n}
\end{eqnarray}
where $\boldsymbol{n}$ is the noise in each pixel and $\boldsymbol{P}$ is
the blurring operator written in matrix form.

The prior for the pixel intensities $\{m_i\}$ has an ``entropic'' form,
where images are upweighted (relative to a uniform prior) if the pixel
intensities have a low dynamic range. The prior density is
\begin{eqnarray}
p(\boldsymbol{m} | \lambda) \propto \exp
\left[
-\lambda \sum_{i=1}^n m_i \ln m_i
\right]
\end{eqnarray}
where $\lambda$ is a hyperparameter.
In the 1980s, this prior was thought to have a fundamental status, but this
is no longer the case \citep{massinf}. Instead, we can consider the configurational
entropy $-\sum_{i=1}^n m_i \ln m_i$ as simply a function of an image that may
be relevant somehow, and that we are uncertain about.

The posterior for the underlying image given the data is
\begin{eqnarray}
p(\boldsymbol{m} | \boldsymbol{D}, \lambda)
&=& \frac{p(\boldsymbol{m} | \lambda)p(\boldsymbol{D}|\boldsymbol{m}, \lambda)}
{\int p(\boldsymbol{m}|\lambda)p(\boldsymbol{D}|\boldsymbol{m}, \lambda) \,d^n\boldsymbol{m}}\\
&=& \frac{\exp\left[-\lambda \sum_i m_i\ln m_i
-\frac{1}{2\sigma^2}\left(D_i - \sum_{j} P_{ij}m_{j}\right)^2\right]}
{\int \exp\left[-\lambda \sum_i m_i\ln m_i
-\frac{1}{2\sigma^2}\left(D_i - \sum_{j} P_{ij}m_{j}\right)^2\right]\, d^n\boldsymbol{m}}
\end{eqnarray}

@@

Prior to knowing the data $D$, we don't know how much information we will get
about the question of $M_1$ vs. $M_2$, quantified by the KL divergence:
\begin{eqnarray}
\mathcal{H} &=& \sum_{i=1}^2 P(M_i | D)
\ln\left[
\frac{P(M_i | D)}{P(M_i)}
\right].
\end{eqnarray}

The expected value of $\mathcal{H}$ (with respect to the prior over datasets)
is the mutual information:
\begin{eqnarray}
\mathcal{I} &=& \mathds{E}\left[\mathcal{H}\right]\\
&=&
\sum_{i=1}^2 \int P(M_i, D)
\ln\left[
\frac{P(M_i, D)}{P(M_i)P(D)}
\right] \, dD.
\end{eqnarray}

>>>>

$Z$ is useful because it is the probability of the data given that
$\xx$ is {\em one of}
the possibilities in the space being integrated over, and thus plays the role
of a likelihood for the parameter space as a whole.
The main goal of NS is to approximate $Z$, but it can also be used to make
Monte Carlo approximations of the posterior distribution, and to
calculate the `information',
or Kullback-Leibler divergence from the prior to the posterior:
\begin{eqnarray}
H &=& \int p(\xx | \data, \II) \ln
\left[\frac{p(\xx | \data, \II)}{p(\xx | \II)}\right] \, d\xx \\
&=& \int \frac{\pi(\xx) L(\xx)}{Z} \ln
\left[\frac{L(\xx)}{Z}\right] \, d\xx.
\end{eqnarray}
$H$ quantifies how compressed the posterior distribution is with
respect to the prior. For example, $H = 10$ implies (loosely speaking)
that the posterior occupies a fraction $e^{-10}$ of the prior mass.
$H$ can be interpreted quite literally as a measure of how much
the data $\data$ told you about $\xx$.

@@

The prior $\pi(\xx)$ implies a certain prior for $L_1$ and $L_2$, which we
denote $\pi(L_1, L_2)$. An example of a prior is shown as the
probability density in Figure~\ref{fig:joint1}. The partition function
$Z(\beta_1, \beta_2)$ is a set of expected values with respect to this density.
Simple Monte Carlo sampling from $\pi$ will not work except for values of
$(\beta_1, \beta_2)$ where the canonical distribution remains similar to $\pi$.
We need a sampler that explores regions where $L_1$ and $L_2$ are high in order
to accurately estimate $Z(\beta_1, \beta_2)$.
\begin{figure}[ht!]
\centering
\includegraphics[scale=0.5]{figures/joint1.pdf}
\caption{\it An example of a prior for $L_1$ and $L_2$, which is implied by
the prior $\pi(\xx)$. We have overplotted 25 points drawn from this prior.
The estimated prior mass in the shaded rectangle is
$\hat{X}(2, 1.5) = 5/25 = 0.2$.
\label{fig:joint1}}
\end{figure}

@@

\begin{figure}[ht!]
\centering
\includegraphics[scale=0.5]{figures/contours.pdf}
\caption{\it The joint prior for the objective functions, and fifteen
contours of the cumulative distribution $X(L_1, L_2)$, each enclosing half
the prior mass of the previous one. \label{fig:contours}}
\end{figure}

@@

\begin{figure}[ht!]
\centering
\includegraphics[scale=0.5]{figures/joint2.pdf}
\caption{\it The joint prior for the objective functions, and Monte Carlo
samples, as in Figure~\ref{fig:joint1}. The shaded region imposes a
constraint on the value of $\hat{X}$. Its boundary approximates a contour
of $X(L_1, L_2)$.
\label{fig:joint2}}
\end{figure}

@@

\begin{figure}[ht!]
\centering
\includegraphics[scale=0.75]{figures/truth.pdf}
\caption{The true partition function and KL divergence (from the prior to
the canonical distribution) as a function of two temperatures for the
test problem.\label{fig:truth}}
\end{figure}

\begin{figure}[ht!]
\centering
\includegraphics[scale=1]{figures/output.png}
\caption{The sequence of discarded particles (yellow) and
points representing the canonical distribution at temperatures
$(T_1, T_2) = (0.1, 1)$ (black).\label{fig:output}}
\end{figure}

\begin{figure}[ht!]
\centering
\includegraphics[scale=0.5]{figures/results.pdf}
\caption{Numerical results for the test problem. These compare well with the
true partition function and KL divergence given in Figure~\ref{fig:truth}.
\label{fig:results}}
\end{figure}

@@

\section{Example 3: ``MaxEnt'' image deconvolution example}

Problems with more than one objective function can also arise in data
analysis. Consider, for example, the problem of deblurring a noisy
image. A common approach to these problems is regularized inversion,
such as the ``maximum entropy'' technique of \citet{gull}. It is important
to distinguish between this method and the maximum entropy principle used
to update or assign probabilities \citep{caticha}.

The unknown parameters are a set of pixel intensities
$\{x_i\}_{i=1}^n$ where $n$ is the number of pixels and $x_i \geq 0$.
The dataset is also a collection of pixel intensities $\{D_i\}_{i=1}^n$.
The prior for the dataset given the parameters is a normal distribution
with mean given by the underlying image $\{xi\}$ convolved by
a point spread function. The standard deviation of each pixel is
$\sigma$, assumed known. This is usually written as

\begin{eqnarray}
\boldsymbol{D} = \boldsymbol{M}\boldsymbol{x} + \boldsymbol{n}
\end{eqnarray}
where $\boldsymbol{n}$ is the noise in each pixel and $\boldsymbol{M}$ is
the blurring operator written in matrix form.

The prior for the pixel intensities $\{x_i\}$ has an ``entropic'' form,
where images are upweighted (relative to a uniform prior) if the pixel
intensities have a low dynamic range. The prior density is
\begin{eqnarray}
p(\boldsymbol{x} | \lambda) \propto \exp
\left[
-\lambda \sum_{i=1}^n x_i \ln x_i
\right]
\end{eqnarray}
where $\lambda$ is a hyperparameter.
In the 1980s, this prior was thought to have a fundamental status, but this
is no longer the case \citep{massinf}. Instead, we can consider the configurational
entropy $-\sum_{i=1}^n x_i \ln x_i$ as simply a function of an image that may
be relevant somehow, and that we are uncertain about.

The posterior for the underlying image given the data is
\begin{eqnarray}
p(\boldsymbol{x} | \boldsymbol{D}, \lambda)
&=& \frac{p(\boldsymbol{x} | \lambda)p(\boldsymbol{D}|\boldsymbol{x}, \lambda)}
{\int p(\boldsymbol{x}|\lambda)p(\boldsymbol{D}|\boldsymbol{x}, \lambda) \,d^n\boldsymbol{m}}\\
&=& \frac{\exp\left[-\lambda \sum_i x_i\ln x_i
-\frac{1}{2\sigma^2}\left(D_i - \sum_{j} P_{ij}x_{j}\right)^2\right]}
{\int \exp\left[-\lambda \sum_i x_i\ln x_i
-\frac{1}{2\sigma^2}\left(D_i - \sum_{j} P_{ij}x_{j}\right)^2\right]\, d^n\boldsymbol{m}}
\end{eqnarray}

@@

Prior to knowing the data $D$, we don't know how much information we will get
about the question of $M_1$ vs. $M_2$, quantified by the KL divergence:
\begin{eqnarray}
H &=& \sum_{i=1}^2 P(M_i | D)
\ln\left[
\frac{P(M_i | D)}{P(M_i)}
\right].
\end{eqnarray}

The expected value of $H$ (with respect to the prior over datasets)
is the mutual information:
\begin{eqnarray}
\mathcal{I} &=& \mathds{E}\left[H\right]\\
&=&
\sum_{i=1}^2 \int P(M_i, D)
\ln\left[
\frac{P(M_i, D)}{P(M_i)P(D)}
\right] \, dD.
\end{eqnarray}
